Completing correlation matrices

### 2021-10-25 paper/paper_risk.tex
Added introduction, abstract, and conclusion.

\documentclass[12pt, oneside]{article} 
\usepackage[a4paper]{geometry}              
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[table]{xcolor}

%SetFonts
\usepackage[T1]{fontenc}
\usepackage[bitstream-charter]{mathdesign}
%SetFonts

%define example environment 
\newcounter{examplecounter}
\newenvironment{example}%
{%
\small\begin{quote}%
\refstepcounter{examplecounter}%
\textbf{Example \arabic{examplecounter}}%
\quad%
}%
{%\schluss%
\end{quote}%
}

%define remark environment 
\newcounter{remarkcounter}
\newenvironment{remark}%
{%
\small\begin{quote}%
\refstepcounter{remarkcounter}%
\textbf{Remark \arabic{remarkcounter}}%
\quad%
}%
{%\schluss%
\end{quote}%
}

%define remark environment 
\newcounter{questioncounter}
\newenvironment{question}%
{%
\small\begin{quote}%
\refstepcounter{questioncounter}%
\textbf{Question \arabic{questioncounter}}%
\quad%
}%
{%\schluss%
\end{quote}%
}

%define important environment
\newenvironment{important}{\begin{quote}%
\textbf{Important:}%
\quad
}{%
\end{quote}%
}

\newcommand{\qed}{\nobreak \ifvmode \relax \else
      \ifdim\lastskip<1.5em \hskip-\lastskip
      \hskip1.5em plus0em minus0.5em \fi \nobreak
      \vrule height0.5em width0.5em depth0.25em\fi}

@@

\usepackage{lettrine}

\begin{document}
\noindent{\Huge \textbf\textsf{{Completing correlation matrices}}}

\noindent \textit{by Horst Köhler, Thomas Streuer, Olaf Dreyer}

\vspace{.5cm}

\section{Introduction}

\section{The completion criterium}
Given a correlation matrix with unknown coefficients what is the right way to complete it? To motivate the answer we need to introduce some notation. Let 
\begin{equation}
	X = ( x_1, \ldots, x_k )^T\in \R^k
\end{equation}
be a $k$ dimensional random variable with a density function given by a $k$ dimensional normal distribution
\begin{align}
	f(X) & = N_k[\mu, H_X]( X ) \\
	& = (2\pi)^{-k/2}\det(H_X)^{-1/2}\exp\left(-\frac{1}{2}(X -\mu)^T H_X^{-1} (X-\mu)\right),
\end{align}
with mean $\mu$ and covariance matrix $H_X$. A crucial property of the normal distribution is that if we condition on some of the $x_i$, $i=1, \dots, k$, we again obtain a normal distribution. Let us assume that we want to condition on the last $l$ components of $X$:
\begin{equation}
	Z = (x_{k-l+1}, \ldots, x_k)^T \in \R^l.
\end{equation}
We partition the matrix $H$ in a way that reflects this partition of $X$:
\begin{equation}
	H_X = \begin{pmatrix}
		A & B \\
		B^T & C
	\end{pmatrix},
\end{equation}
with $A\in M_{k-l}$, $C\in M_l$, and $B\in M_{k-l,l}$. Let $\bar X$ be the first $k-l$ components of $X$:
\begin{equation}
	\bar X = (x_1, \ldots, x_{k-l})^T\in \R^{k-l}
\end{equation}
If we assume that $X$ has zero mean (i.e. $\mu=0$), the density function for $X$ conditioned on $Z$ is then given by
\begin{equation}
	f(X \vert Z ) = N_{k-l}[BC^{-1}Z, H_X/C ](\bar X),
\end{equation}
(see \cite[chapter 8]{rao} and \cite{cottle}) where $H_X/C$ is the Schur complement of $C$ in $H_X$:
\begin{equation}\label{eqn.hxc}
	H_X/C = A - B C^{-1}B^T
\end{equation}
(Emilie Haynsworth introduced the name and highlighted its usefulness in \cite{hayns}. Schur originally made use of it in \cite{schur}. For an overview of the properties of the Schur complement see \cite{horn}.) We see that we again obtain a normal distribution. We will meet the Schur complement again soon.

@@

\section{The completion procedure}
In the last section we have completed the matrix $H$ in the very simple case where there are two known matrices $H_X$ and $H_Y$ with just one shared part $C$. The general case is usually more complicated. Building on the result in \cite{grone} Smith \cite{smith} (see also \cite{ksd} for a more general treatment of the semidefinite case) showed how to solve the general problem by reducing it to a repeated application of the result from the previous section. We will discuss the required steps in this section by looking at a particular example: the cross currency model.

@@

\section{Several currencies}
In the previous section we have just looked at one foreign currency $A$ with one exchange rate $X$. The procedure can easily be extended to an arbitrary number of foreign currencies. We will denote the foreign currencies by a capital letter:
\begin{equation}
	A, B, C, \ldots
\end{equation}
The exchange rate between the domestic currency $E$ and the foreign currency $A$ will now be denoted by 
\begin{equation}
	X_E^A.
\end{equation}
If $m_A$ is an amount in currency $A$ then $X_E^Am_A$ is the corresponding amount in currency $E$. If we are given the same correlation coefficients as in the previous section we obtain the graph in figure \ref{fig.several}. The corresponding graph of cliques is shown in figure \ref{fig.severalCliques}. The corresponding clique tree is obtained by erasing the lines that connect the cliques 
\begin{equation}
	\alpha_{CA}, \alpha_{CB}, \ldots, \alpha_{CG}
\end{equation}
on the middle layer. The intersection of any two of these cliques is given by
\begin{equation}
	\{E\} \subset \alpha_E.
\end{equation}
Since the path connecting the two cliques passes through $\alpha_E$ this tree satisfies the intersection property.

@@

\begin{figure}[hbt]
  \begin{center}
  	\includegraphics[width=12cm]{img/severalCurrCliques.png}
  \end{center}
  \caption{The graph of cliques for the case of several foreign currencies. A clique tree is obtained by removing the lighter edges in the middle layer. It is again a tree of height two.}\label{fig.several}
\end{figure}

\section{Conclusion}

\begin{thebibliography}{MM}
	\bibitem{rao} C. Radharkishna Rao, Linear Statistical Inference and its Applications, 2nd edition, John Wiley \& Sons, Inc, 2002.
	\bibitem{cottle} Richard W. Cottle, Manifestations of the Schur complement, Linear Algebra and its Applications \textbf{8}, 1974, 189--211. 
	\bibitem{hayns} Emilie Virginia Haynsworth, Determination of the inertia of a partitioned Hermitian matrix,  Linear  Algebra and its Applications \textbf{1}, 1968, 73--81.
	\bibitem{schur} J. Schur, Über Potenzreihen, die im Innern des Einheitskreises beschränkt sind, Journal für die reine und angewandte Mathematik \textbf{147}, 205 -- 232, 1917.
	\bibitem{horn} Roger A. Horn, Basic  Properties  of  the  Schur  Complement. In Fuzhen Zhang (Ed.) The Schur complement and its application (17--46). Springer, 2005.
	\bibitem{grone} R. Grone, C.R. Johnson, E. Sa, H. Wolkowicz, Positive definite completion of partial Hermitian matrices, Linear Algebra Appl. 58, 109--124, 1984.
	\bibitem{smith} R. L. Smith, The positive definite completion problem revisited, Linear Algebra and its Applications 429 (2008) 1442 -- 1452.
	\bibitem{ksd} our other paper ...
	\bibitem{golumbic} M. C. Golumbic, Algorithmic Graph Theory and Perfect Graphs, Elsevier, 2004. 
	\bibitem{blair} Jean R. S. Blair, Barry Peyton, An introduction to chordal graphs and clique trees, in Alan George, John R. Gilbert, Joseph W. H. Liu (Eds.), Graph Theory and Sparse Matrix Computation (pp. 1 -- 29), The IMA Volumes in Mathematics and its Applications, 56, Springer Verlag, 1993.
\end{thebibliography}

>>>>

\documentclass[11pt, a4paper]{article} 
% \usepackage[a4paper]{geometry}              
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,  % hyperlinks will be coloured
    linkcolor=blue,   % hyperlink text will be blue
    citecolor=blue
}
\usepackage{xcolor}
% \usepackage{a4wide}

@@

\begin{document}
\title{Completing correlation matrices}

\author{Olaf Dreyer\thanks{OD Consulting}
\and Horst K\"ohler \thanks{Commerzbank AG}
\and Thomas Streuer \thanks{RIVACON GmbH} }

\date{\today}

\maketitle

\begin{abstract}
	We describe a way to complete a correlation matrix that is not fully specified. Such matrices arise in financial applications when the number of stochastic variables becomes large or when several smaller models are combined in a larger model. We argue that the proper completion to consider is the matrix that maximizes the entropy of the distribution described by the matrix. We then give a way to construct this matrix starting from the graph associated with the incomplete matrix. If this graph is chordal our construction will result in a proper correlation matrix. We demonstrate the construction using a cross-currency model with six stochastic variables.  
\end{abstract}

\textbf{Keywords:} stochastic models, hybrid models, stochastic volatility, correlation matrices, graph theory, chordal graphs

\textbf{JEL:} C02, C63, G12

\section{Introduction}
Modern financial models are driven by a large number of stochastic variables. These stochastic variables are usually correlated which creates the need to determine the coefficients of the underlying correlation matrix. Since the number of correlation coefficients increases quadratically with the number of stochastic variables, this number can become very large. While it would be desirable to calibrate all these coefficients this is in practice not possible because such a calibration procedure would be too slow, too unstable, or would require calibration instruments for which quotes can not be obtained. Hybrid models are another source of undetermined correlation coefficients. Two calibrated models might be parts of a larger model. While the coefficients for the individual models have been determined in separate calibration, the coefficients that mix variables from the two different models are undetermined. In a large model, with many smaller constituent models, the number of such coefficients can constitute the bulk of the matrix.

Faced with this situation, it is tempting to set the missing coefficients to zero. The idea here is that any other choice seems to add arbitrary information. This approach has several shortcomings though. The most important of which is that one might end up with a matrix that is not positive semidefinite and thus not a valid correlation matrix. To obtain a valid matrix further steps are required. One possibility is to use the fact that the set of positive definite correlation matrices is an open set that contains the unit matrix. It is thus always possible to scale the off-diagonal elements of the matrix with a common scale factor $\lambda<1$ and obtain a positive definite matrix. Another possibility is to choose the nearest correlation matrix in a suitable topology (see e.g \cite{higham}). These additional steps might change all the coefficients of the matrix, including those coefficients, that have been obtained in previous calibrations. They also rely on somewhat arbitrary further assumptions like a particular notion of distance on the space of matrices.

The approach presented here does not suffer from these shortcomings. We start with a particular criterium that singles out one particular completion. The criterium is distinguished by the fact that it puts the least amount of additional constraints on the distribution described by the completed matrix. As it turns out, this is not the matrix that contains zeroes in the places of unknown coefficients. Rather, it is the inverse of the matrix that contains zeroes in those places that were undetermined before. This criterium removes the arbitrary nature of other constructions.

We then show how to construct this completion. The construction starts by looking at the graph associated with the incomplete matrix. The vertices of the graph are the rows (or columns) of the matrix. Since we are looking at a correlation matrix, the vertices correspond to the stochastic variables of the model. Two vertices are connected by an edge in the graph if the corresponding correlation coefficient in the matrix is know. It is here that we encounter the one condition that needs to be satisfied for the construction to work. The graph that we have just constructed needs to be chordal. A graph is chordal if every loop in the graph that has more than three edges contains a chord, i.e. an edge that connects two non-adjacent vertices in the loop. If this condition is satisfied we can follow a construction first introduced by \cite{grone} and refined by \cite{smith} to complete the matrix. Every step of the construction adds new vertices to the edge which in turn correspond to entries in the correlation matrix. The matrix obtained in this way will automatically be a positive-semidefinite so that no further steps need to be taken to obtain a valid correlation matrix.

The paper now consists of three parts. In the next section we discuss the completion criterium. The matrix can be characterized in three different ways and we present them here. We then discuss the construction of the completion in section \ref{sec.procedure}. We do this by following the concrete example of a cross-currency model with two currencies and one exchange rate. The two models for the interest rate and the model for the exchange rate are assumed to be stochastic volatility models so that we end up with a total of six stochastic variables. We describe all the required steps using this model but the steps can easily be generalized to more complicated models. One such model is the $N$-currency model with $(N-1)$ foreign exchange rates that is described in section \ref{sec.several}.

\section{The completion criterium}\label{sec.criterium}
Given a correlation matrix with unknown coefficients what is the right way to complete it? To motivate the answer we need to introduce some notation. Let 
\begin{equation}
	X = ( x_1, \ldots, x_k )^T\in \R^k
\end{equation}
be a $k$ dimensional random variable with a density function given by a $k$ dimensional normal distribution
\begin{align}
	f(X) & = N_k[\mu, H_X]( X ) \\
	& = (2\pi)^{-k/2}\det(H_X)^{-1/2}\exp\left(-\frac{1}{2}(X -\mu)^T H_X^{-1} (X-\mu)\right),
\end{align}
with mean $\mu$ and covariance matrix $H_X$. A crucial property of the normal distribution is that if we condition on some of the $x_i$, $i=1, \dots, k$, we again obtain a normal distribution. Let us assume that we want to condition on the last $l$ components of $X$:
\begin{equation}
	Z = (x_{k-l+1}, \ldots, x_k)^T \in \R^l.
\end{equation}
We partition the matrix $H$ in a way that reflects this partition of $X$:
\begin{equation}
	H_X = \begin{pmatrix}
		A & B \\
		B^T & C
	\end{pmatrix},
\end{equation}
with $A\in M_{k-l}$, $C\in M_l$, and $B\in M_{k-l,l}$. Let $\bar X$ be the first $k-l$ components of $X$:
\begin{equation}
	\bar X = (x_1, \ldots, x_{k-l})^T\in \R^{k-l}
\end{equation}
If we assume that $X$ has zero mean (i.e. $\mu=0$), the density function for $X$ conditioned on $Z$ is then given by
\begin{equation}
	f(X \vert Z ) = N_{k-l}[BC^{-1}Z, H_X/C ](\bar X),
\end{equation}
(see \cite[chapter 8]{rao} and \cite{cottle}) where $H_X/C$ is the Schur complement of $C$ in $H_X$:
\begin{equation}\label{eqn.hxc}
	H_X/C = A - B C^{-1}B^T
\end{equation}
(Emilie Haynsworth introduced the name and highlighted its usefulness in \cite{hayns}. Schur originally made use of it in \cite{schur}. For an overview of the properties of the Schur complement see \cite{horn}.) We see that we again obtain a normal distribution. We will meet the Schur complement again soon.

@@

\section{The completion procedure}\label{sec.procedure}
In the last section we have completed the matrix $H$ in the very simple case where there are two known matrices $H_X$ and $H_Y$ with just one shared part $C$. The general case is usually more complicated. Building on the result in \cite{grone} Smith \cite{smith} (see also \cite{ksd} for a more general treatment of the semidefinite case) showed how to solve the general problem by reducing it to a repeated application of the result from the previous section. We will discuss the required steps in this section by looking at a particular example: the cross currency model.

@@

\section{Several currencies}\label{sec.several}
In the previous section we have just looked at one foreign currency $A$ with one exchange rate $X$. The procedure can easily be extended to an arbitrary number of foreign currencies. We will denote the foreign currencies by a capital letter:
\begin{equation}
	A, B, C, \ldots
\end{equation}
The exchange rate between the domestic currency $E$ and the foreign currency $A$ will now be denoted by 
\begin{equation}
	X_E^A.
\end{equation}
If $m_A$ is an amount in currency $A$ then $X_E^Am_A$ is the corresponding amount in currency $E$. If we are given the same correlation coefficients as in the previous section we obtain the graph in figure \ref{fig.several}. The corresponding graph of cliques is shown in figure \ref{fig.severalCliques}. The corresponding clique tree is obtained by erasing the lines that connect the cliques 
\begin{equation}
	\alpha_{CA}, \alpha_{CB}, \ldots, \alpha_{CG}
\end{equation}
on the middle layer. The intersection of any two of these cliques is given by
\begin{equation}
	\{E\} \subset \alpha_E.
\end{equation}
Since the path connecting the two cliques passes through $\alpha_E$ this tree satisfies the intersection property.

@@

\begin{figure}[hbt]
  \begin{center}
  	\includegraphics[width=12cm]{img/severalCurrCliques.png}
  \end{center}
  \caption{The graph of cliques for the case of several foreign currencies. A clique tree is obtained by removing the lighter edges in the middle layer. It is again a tree of height two.}\label{fig.severalCliques}
\end{figure}

\section{Conclusion}\label{sec.conclusion}
In this paper we have described a method to complete an insufficiently specified correlation matrix. Because financial models are getting more complex such matrices are a common occurrence.

\section*{Acknowledgement}
The authors would like to thank Patrick B\"uchel for his support during the creation of this work; and Nataliya Koval, and Roland Seydel for insightful discussions on the subject.

\begin{thebibliography}{MM}
	\bibitem[Higham, 2002]{higham}, Higham, N. J. (2002). Computing the nearest correlation matrix - a problem from finance. \textit{IMA Journal of Numerical Analysis} 22: 329 - 343.
		\bibitem[Grone et~al., 1984]{grone} Grone R., Johnson C.R., Sa E., Wolkowicz H. (1984). Positive definite completion of partial Hermitian matrices. \textit{Linear Algebra Appl.} 58: 109--124.
	\bibitem[Smith, 2008]{smith} Smith, R. L. (2008). The positive definite completion problem revisited. \textit{Linear Algebra and its Applications} 429:1442 -- 1452.
	\bibitem[Rao, 2002]{rao} C. R. Rao (2002). \textit{Linear Statistical Inference and its Applications, 2nd edition}. John Wiley \& Sons, Inc.
	\bibitem[Cottle, 1974]{cottle} Cottle, R. W. (1974). Manifestations of the Schur complement. \textit{Linear Algebra and its Applications} 8:189--211. 
	\bibitem[Haynsworth, 1968]{hayns} Haynsworth, E.V. (1968), Determination of the inertia of a partitioned Hermitian matrix,  \textit{Linear  Algebra and its Applications} 1:73--81.
	\bibitem[Schur, 1917]{schur} Schur, J. (1917). \"Uber Potenzreihen, die im Innern des Einheitskreises beschr\"ankt sind. \textit{Journal f\"ur die reine und angewandte Mathematik} 147: 205--232.
	\bibitem[Horn, 2005]{horn} Horn, R. A. (2005). Basic  Properties  of  the  Schur  Complement, 17--46. In Fuzhen Zhang (Ed.) \textit{The Schur complement and its application}. Springer.
	\bibitem[Dreyer et~al., 2021] {ksd} our other paper ...
	\bibitem[Golumbic, 2004]{golumbic} Golumbic, M. C. (2004). \textit{Algorithmic Graph Theory and Perfect Graphs}. Elsevier. 
	\bibitem[Blair et~al., 1993]{blair} Blair, J. R. S., Peyton, B. (1993). An introduction to chordal graphs and clique trees, 1 -- 29. In George, A., Gilbert, J. R., Liu, J. W. H. (Eds.). \textit{Graph Theory and Sparse Matrix Computation}, The IMA Volumes in Mathematics and its Applications 56. Springer.
\end{thebibliography}
